---
title: "The ARS chain: from an existing programme to a study-level ARD"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
```

The CDISC **Analysis Results Standard (ARS)** describes, as metadata, every
analysis a reporting event needs -- populations, groupings, data subsets,
statistical methods, and the displays they feed. Once that metadata exists,
the pharmaverse package [siera](https://cran.r-project.org/package=siera)
turns it into runnable [cards](https://insightsengineering.github.io/cards/)
programmes that produce **Analysis Results Datasets (ARDs)**.

The catch is authoring the ARS itself: the workbook `siera::readARS()`
consumes needs nine interlinked sheets with heavily repeated information.
ydisctools closes that gap with a small toolchain, and this article walks the
**whole chain** on runnable code:

```
existing programme (cards / cardx)     existing ARD
        |  ars_params_from_code()              |  ars_params_from_ard()
        +----- ars_params_recover() ... merge both, review the notes ----+
        v
compact parameter workbook  (1 row per analysis)
        |  build_ars() + write_ars_xlsx()  ... expand to the full ARS metadata
        v
ARS workbook  (siera-ready)
        |  ars_generate_ard()              ... siera::readARS() + run + combine
        v
one study-level ARD, keyed by OutputId
```

```{r setup, message = FALSE}
library(ydisctools)
```

Everything below runs on the example set shipped with the package -- dummy
ADaM data (30 subjects, three CDISC-pilot arms) and the artefacts of a
Demographics and an Adverse Events display:

```{r}
ex <- system.file("ars-examples", package = "ydisctools")
list.files(ex, recursive = TRUE)
```

## 1. The compact parameter workbook

The chain's central object is a small workbook: one `Outputs` row per
display, one `Analyses` row per analysis, and (optionally) a `Displays`
sheet for titles, footnotes and page furniture. `ars_param_template()`
writes a filled-in example you can copy per study:

```{r}
params_file <- file.path(tempdir(), "params.xlsx")
ars_param_template(params_file, overwrite = TRUE)
params <- read_ars_params(params_file)
params$analyses[, c("output_id", "analysis_id", "method", "dataset",
                    "variable", "group_by", "where", "denominator")]
```

Three conventions do most of the work:

- **Conditions** use a mini syntax: `"SAFFL"` is sugar for
  `ADSL.SAFFL EQ Y`, `"TRTEMFL EQ Y"` is a full condition, `";"` chains
  conditions with AND, and `IN` values are pipe-separated
  (`"SEX IN M|F"`).
- **"n (%) of subjects per category"** keeps `variable = USUBJID` and puts
  the category variable (AGEGR1, SEX, AEBODSYS, ...) in as the *second*
  grouping -- the ARS idiom.
- **`denominator = "auto"`** wires a percentage analysis to the `total_n`
  analysis with the same population and first grouping in the same output.

You can author this workbook by hand. But if a similar study already has
programmes, you do not have to start from a blank sheet -- that is step 2.

## 2. Recover parameters from an existing programme

`ars_params_from_code()` statically analyses R programmes (nothing is
executed) and drafts the parameter set from the {cards} / {cardx} calls it
finds. Here is a compact "existing study programme" in the style of the
[rtfreporter showcase](https://ichirio.github.io/rtfreporter/articles/showcase-dm.html):

```{r}
old_programme <- file.path(tempdir(), "t_dm_old_study.R")
writeLines('
library(cards); library(dplyr)
arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
adsl <- pharmaverseadam::adsl |>
  filter(TRT01A %in% arm_levels, SAFFL == "Y")

dm_header <- rtf_header(rows = list(
  c(l = "Acme Biopharma, Inc.", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
  c(c = "Table 14.1.1"),
  c(c = "Summary of Demographic Data")))
dm_footer <- rtf_footer(rows = list(
  c(l = "Note: Percentages are based on the safety population."),
  c(l = "Source: ADSL", r = "CONFIDENTIAL")))

ard <- ard_stack(
  data = adsl, .by = TRT01A,
  ard_continuous(variables = AGE),
  ard_categorical(variables = c(AGEGR1, SEX, RACE)),
  .total_n = TRUE)
', old_programme)

rec <- ars_params_from_code(old_programme,
                            output_id = "Out_dm",
                            output_name = "Summary of Demographic Data")
rec$analyses[, c("analysis_id", "method", "dataset", "variable",
                 "population", "group_by", "groups", "denominator")]
```

The `filter()` pipeline was classified, not just copied: the `%in%` on the
grouping variable became the pre-defined **group levels**, and the `SAFFL`
flag became the **population**. The display furniture was captured too:

```{r}
rec$displays
```

Everything the static analysis could not decide is an explicit note --
review these before moving on (they also land on a `Notes` sheet in the
written workbook):

```{r}
rec$notes
```

Save the draft, edit what the notes flag, and it is a normal parameter
workbook:

```{r}
draft_file <- file.path(tempdir(), "params_recovered.xlsx")
write_ars_params(rec, draft_file, overwrite = TRUE)
readxl::excel_sheets(draft_file)
```

This also works on siera-generated ARD programmes -- recovering the shipped
`ARD_Out_dm.R` reproduces the parameter set it was built from:

```{r}
ars_params_from_code(file.path(ex, "ARD_Out_dm.R"),
                     output_id = "Out_dm")$analyses[
  , c("method", "dataset", "variable", "population", "group_by")]
```

## 3. No code at hand? Start from the ARD -- or use both

Sometimes the programme is gone, was written in another language, or simply
cannot be parsed -- but its **ARD survives**.  `ars_params_from_ard()`
drafts the same parameter set from the ARD itself (a cards `ard_stack()`
result, a siera-stamped combined ARD, or a flattened csv/xlsx export).
Here it reads the ten-display study ARD bundled with the
[SAP-to-tables pipeline](sap-to-ars.html):

```{r}
ard_file <- file.path(system.file("sap-pipeline", package = "ydisctools"),
                      "05_ard", "ARD_STUDY01.csv")
rec_ard <- ars_params_from_ard(ard_file)
rec_ard$analyses[rec_ard$analyses$output_id == "Out_dm",
                 c("name", "method", "variable", "group_by", "denominator")]
```

An ARD is authoritative about **which analyses ran**, but it carries no
data provenance: population, subsets and dataset names come back blank with
REVIEW notes.  The programme code is exactly the opposite -- it knows the
provenance but may be blind to a variable list it cannot resolve.  The two
sources are complementary, and `ars_params_recover()` merges them.
Consider a programme whose variable lists live in a `source()`d setup file
this analysis cannot see:

```{r}
src <- file.path(tempdir(), "t_dm_indirect.R")
writeLines('
library(cards); library(dplyr)
source("study_setup.R")     # defines vars_cont / vars_cat -- unreadable here
arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
adsl <- ADSL |> filter(TRT01A %in% arm_levels, SAFFL == "Y")
ard <- adsl |>
  ard_stack(ard_summary(variables = all_of(vars_cont)),
            ard_tabulate(variables = all_of(vars_cat)),
            .by = TRT01A, .total_n = TRUE)', src)
```

Alone, the static analysis would keep the unresolved symbols (with REVIEW
notes).  Handed the ARD as well, the merge takes the analysis list from the
ARD and the provenance from the code:

```{r}
ard_dm <- read.csv(ard_file, stringsAsFactors = FALSE)
ard_dm <- ard_dm[ard_dm$OutputId == "Out_dm", ]

rec <- ars_params_recover(src, ard_dm)
rec$analyses[, c("name", "method", "variable", "dataset", "population")]
rec$outputs$groups
```

Every real variable came from the ARD; `ADSL`, `SAFFL` and the pre-defined
arm levels came from the code.  Disagreements between the two sources --
analyses only in the code (conditional? superseded?) or only in the ARD --
are reported as explicit notes, so the cross-check is itself part of the
review:

```{r}
grep("no counterpart|not in the programme code", rec$notes, value = TRUE)
```

## 4. Expand to the full ARS workbook

`build_ars()` does the tedious part: de-duplicated analysis sets, groupings
and data subsets with automatic IDs, numerator/denominator wiring, contents
lists, method sheets (expanded from siera's own MIT-licensed method-template
catalog, so nobody writes template code), and -- because our recovered
parameters carry display metadata -- the ARS `Outputs` / `Displays` /
`GlobalDisplaySections` sheets:

```{r}
ars <- build_ars(params_file)   # the template example from step 1
names(ars)
ars$AnalysisSets
head(ars$Analyses[, c("id", "analysisSetId", "groupingId1", "groupingId2",
                      "dataSubsetId", "method_id")])
```

`write_ars_xlsx()` writes it as a workbook `siera::readARS()` accepts
directly (for the JSON ARS representation, convert the written workbook with
`siera::ars_xlsx_to_json()`, available in siera > 0.5.6):

```{r}
ars_file <- file.path(tempdir(), "ARS_STUDY01.xlsx")
write_ars_xlsx(ars, ars_file, overwrite = TRUE)
```

## 5. Generate, run, combine: one ARD for the study

`ars_generate_ard()` finishes the chain. It hands the ARS to
`siera::readARS()` (one generated programme per output, kept for
inspection), runs each programme in an isolated environment against your
ADaM folder, and binds the results. siera stamps `AnalysisId` / `MethodId` /
`OutputId` onto every ARD row, so the combined frame is keyed by display out
of the box:

```{r, message = FALSE, warning = FALSE}
ard <- ars_generate_ard(ars_file, adam_path = file.path(ex, "adam"))
table(ard$OutputId)
ard[ard$AnalysisId == "An_03" & ard$stat_name %in% c("n", "p"),
    c("OutputId", "AnalysisId", "group1_level", "variable",
      "variable_level", "stat_name", "stat")] |> head(8)
```

Note how readable the categorical rows are: the treatment arm sits in
`group1_level` and the category variable keeps its real name and levels
(`variable = "AGEGR1"`, `variable_level = "<65"`, ...) -- ydisctools ships an
improved categorical method template (an overlay over siera's catalog) that
uses the modern cards `by = ` / `variables = ` pattern instead of the legacy
dummy-variable idiom.

The generated programmes stay available for review and re-execution:

```{r}
basename(attr(ard, "scripts"))
```

And because `ars` can also be the sheet list itself, the whole chain is one
pipe:

```{r, eval = FALSE}
ard <- read_ars_params("params.xlsx") |>
  build_ars() |>
  ars_generate_ard(adam_path = "path/to/adam")
```

## 6. Where this fits

- **Upstream**, the parameter workbook is the study's analysis intent -- and
  it can be derived from the SAP itself: [From the SAP to the
  tables](sap-to-ars.html) walks that full chain (SAP -> TOC -> shells ->
  ARS -> ARD -> tables) on a bundled ten-display sample study.
  (3+ simultaneous groupings for nested SOC/PT displays are still tracked in
  [ydisctools#6](https://github.com/ichirio/ydisctools/issues/6), currently
  flattened with a note.)
- **Downstream**, the ARD is framework-neutral: shape it with ydisctools'
  `format_stats()` / `pivot_stats_wider()` (the bundled
  `sap-pipeline/06_tfl/make_tfl.R` table programme shows the pattern) or
  `gtsummary::tbl_ard_summary()`, and hand the result to
  [rtfreporter](https://ichirio.github.io/rtfreporter/) for the final RTF --
  or export CDISC Dataset-JSON via siera's `output_format = "datasetjson"`.
- The full worked artefact set of THIS article (parameters, ARS, generated
  programmes, dummy ADaM) ships in
  `system.file("ars-examples", package = "ydisctools")`; the SAP article's
  ten-display study, with every stage artefact through the final RTF, ships
  in `system.file("sap-pipeline", package = "ydisctools")` -- see their
  READMEs.
