---
title: "Building a clinical listing"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
has_rtf <- requireNamespace("rtfreporter", quietly = TRUE)
```

A **listing** prints the data itself, one record per subject, rather than a
summary of it. That sounds simpler than a table -- until you notice that a
listing record rarely fits on one line:

```
Unique       Primary Diagnosis/       Stage at
Subject ID   Any (BRCA) Mutations/    Initial
             Histology                Diagnosis
------------------------------------------------
63016-204    Ovarian cancer/          IIIC
             BRCA1 mutation/
             Serous adenocarcinoma
```

One logical record occupies four physical rows, its cells have to break at
sensible places, every column of the record has to stay vertically aligned,
and a page break must never fall *inside* a record. Doing that by hand means
a pile of string surgery before you ever get to the report.

`rtf_listing()` does that shaping for you. It is a **preparation** step: it
turns a tidy data frame into a listing-ready table plus the metadata that
describes it, and hands the actual RTF off to
[rtfreporter](https://ichirio.github.io/rtfreporter/).

```
tidy ADSL
   |  listing_col()  ... declare each display column
   |  rtf_listing()  ... compose / wrap / align / expand
   v
rtflisting  (data + col_header + col_rel_width + col_spec + hidden record key)
   |  listing_to_rtftables()
   v
rtftable pages  ->  rtfreporter::rtf_tables()  ->  RTF
```

```{r setup}
library(ydisctools)
library(dplyr)
```

## The sample data

A small synthetic baseline-characteristics ADSL, shaped like the oncology
data a listing of this kind is usually built from. Nothing here is real.

```{r sample-data}
adsl <- tibble::tribble(
  ~USUBJID,    ~DISPTPD,                    ~BRCA,             ~HIST,
  "63016-201", "Ovarian cancer",            "BRCA1 mutation",  "Serous adenocarcinoma",
  "63016-202", "Fallopian tube cancer",     "BRCA2 mutation",  "Endometrioid carcinoma",
  "63016-203", "Primary peritoneal cancer", NA,                "Clear cell carcinoma",
  "63016-204", "Ovarian cancer",            "No mutation",     "Serous adenocarcinoma",
  "63016-205", "Ovarian cancer",            "BRCA1 mutation",  "Mucinous carcinoma"
) %>%
  mutate(
    INIDGCAT = c("12.5", "38.1", "6.2", "24.9", "51.3"),
    STAGE    = c("IIIC", "IV", "IIIB", "IIIC", "IIIA"),
    HISTGRD  = c("Grade 3", "Grade 2", "Grade 3", "Grade 1", "Grade 2"),
    PRRAD    = c("Yes", "No", "No", "Yes", "No"),
    PRANTNM2 = c("2 lines", "1 line", "3 lines", "2 lines", "1 line"),
    CMBRFST  = c("Partial response", "Complete response", "Stable disease",
                 "Partial response", "Progressive disease"),
    CMBRLST  = c("Stable disease", "Partial response", "Progressive disease",
                 "Complete response", "Stable disease"),
    PRSRG    = c("Yes", "Yes", "No", "Yes", "No"),
    PPLATFI  = c(8.4321, 14.2087, 3.1234, 21.9876, 5.5555),
    PLATFI   = c(6.1234, 11.8765, 2.4321, 18.5432, 4.3210),
    ECOGPS   = c("0", "1", "1", "0", "2"),
    FOLREVAL = c("2+", "1+", "3+", "2+", "1+")
  )

glimpse(adsl)
```

`listing_col()` takes **column names, not expressions**, so any value you need
to compute -- a rounding, a format, a derived flag -- becomes a column first:

```{r derive}
adsl <- adsl %>%
  mutate(
    PPLATFI_R = round(PPLATFI, 2),
    PLATFI_R  = round(PLATFI,  2)
  )
```

## Choosing the column widths

`width` is the heart of the layout. It is a **display width in characters**
(a full-width Japanese glyph counts as two) and it does double duty: cells
wider than it are wrapped, and the widths together become the relative column
widths the renderer uses.

Picking ten of those by hand is the tedious part, so start from a proposal.
`suggest_listing_widths()` measures each column's real content, floors it by
the column's own header, and scales everything to the width budget you have:

```{r suggest-widths}
suggest_listing_widths(
  adsl,
  listing_col(USUBJID,                    header = "Unique\nSubject ID"),
  listing_col(DISPTPD, BRCA, HIST,        header = "Primary Diagnosis/\nBRCA/\nHistology"),
  listing_col(INIDGCAT,                   header = "Months since\nInitial Diagnosis"),
  listing_col(STAGE,                      header = "Stage at\nInitial\nDiagnosis"),
  listing_col(HISTGRD, PRRAD, PRANTNM2,   header = "Histologic Grade/\nPrior Radiation/\nPrior Lines"),
  listing_col(CMBRFST, CMBRLST,           header = "Best Response\n1st Line/\nLast Line"),
  listing_col(PRSRG,                      header = "Prior\nSurgery"),
  listing_col(PPLATFI_R, PLATFI_R,        header = "Primary Platinum-free/\nPlatinum-free\nInterval (months)"),
  listing_col(ECOGPS,                     header = "ECOG\nPS"),
  listing_col(FOLREVAL,                   header = "FRa\nScore"),
  total_width = 140
)
```

Treat the result as a **starting point**, not an answer: render the listing,
look at it, and nudge the numbers. A column whose cells keep wrapping onto a
third line wants a few more characters; one with white space to spare can give
some back.

## Building the listing

```{r build}
lst <- rtf_listing(
  adsl,
  listing_col(USUBJID,
              header = "Unique\nSubject ID",                             width = 12),
  listing_col(DISPTPD, BRCA, HIST,
              header = "Primary Diagnosis/\nBRCA/\nHistology",           width = 24),
  listing_col(INIDGCAT,
              header = "Months since\nInitial Diagnosis",                width = 10),
  listing_col(STAGE,
              header = "Stage at\nInitial\nDiagnosis",                   width = 9),
  listing_col(HISTGRD, PRRAD, PRANTNM2,
              header = "Histologic Grade/\nPrior Radiation/\nPrior Lines", width = 18),
  listing_col(CMBRFST, CMBRLST,
              header = "Best Response\n1st Line/\nLast Line",            width = 20),
  listing_col(PRSRG,
              header = "Prior\nSurgery",                                 width = 8),
  listing_col(PPLATFI_R, PLATFI_R,
              header = "Primary Platinum-free/\nPlatinum-free\nInterval (months)", width = 16),
  listing_col(ECOGPS,
              header = "ECOG\nPS",                                       width = 6),
  listing_col(FOLREVAL,
              header = "FRa\nScore",                                     width = 7)
)

lst
```

The print-out is the quickest feedback loop you have: five source records
became the physical rows shown, and you can see straight away where the text
broke.

## What the engine did

Four things, in order.

**Composed** each display column from its source columns, joining with the
column's `sep` (`"/"` by default) and skipping `NA` -- the
[`catx()`](../reference/catx.html) convention. Subject `63016-203` has no
`BRCA` value, and the result is
`"Primary peritoneal cancer/Clear cell carcinoma"`, not a dangling `//`:

```{r composed}
lst$data[lst$data[[".record_id"]] == 3L, 3L]
```

**Wrapped** every cell to its column's width, preferring meaningful breaks.
The priority is: after the separator first, then at a word boundary
(whitespace, comma or hyphen), and only then a hard split of a token that is
too long on its own. So a composed diagnosis breaks between its components
rather than mid-word.

**Aligned** each record: every cell is padded to the record's tallest cell, so
a record's lines stay in step across all ten columns. That is why the first
row of a record carries the subject id and the continuation rows are blank in
that column:

```{r record-shape}
rec1 <- lst$data[lst$data[[".record_id"]] == 1L, c(1L, 3L)]
rec1
```

**Expanded and framed** the result: one logical record became several physical
rows, narrow spacer columns were inserted between the content columns, and a
hidden `.record_id` column was added to mark where each record starts and
ends.

```{r structure}
names(lst$data)
lst$col_rel_width
```

The header vector, the relative widths and the per-column alignment are all
derived for you -- including the empty headers and width-1 entries for the
spacers -- so there is no parallel vector to keep in sync by hand:

```{r header}
lst$col_header
```

If you do not want the spacer columns, `spacer_width = 0` leaves them out
entirely (a zero-width column is not a thing the renderer accepts, so `0`
means "none" rather than "invisible"):

```{r no-spacers}
rtf_listing(adsl,
            listing_col(USUBJID, width = 12),
            listing_col(STAGE,   width = 9),
            spacer_width = 0) |>
  (\(x) names(x$data))()
```

## Paginating and rendering

`listing_to_rtftables()` is the single point of contact with rtfreporter. It
passes the hidden record key as both the grouping column and a dropped column,
which lets rtfreporter's `"group_safe"` split **pack whole records onto a page
and never cut one in half** -- the behaviour a listing needs, using the
paginator that already exists rather than a second one:

```{r paginate, eval = has_rtf}
pages <- listing_to_rtftables(lst, max_rows = 12)
length(pages)
```

Each page is an ordinary `rtftable`. The record key is gone, and the
position-indexed metadata has been reindexed to the columns that remain:

```{r page-shape, eval = has_rtf}
names(pages[[1]]$data)
pages[[1]]$col_rel_width
```

Blank rows -- above the first record, between records, and at the foot of the
page -- are carried as positions and materialised by the renderer:

```{r blank-rows, eval = has_rtf}
pages[[1]]$blank_rows
```

From here it is ordinary rtfreporter. Note that the column headers and widths
travel **with** the pages, so `rtf_tables()` needs neither:

```{r render, eval = has_rtf}
library(rtfreporter)

titles <- list(
  c("Listing 16.2.4.2.1.2"),
  c("Baseline Characteristics"),
  c("<Safety Analysis Set>")
)

footers <- list(
  c(l = "ECOG: Eastern Cooperative Oncology Group"),
  c(l = "Note: synthetic data, for illustration only.")
)

rep <- rtf_document(
    page = list(
      orientation     = "landscape",
      paper_size      = "A4",
      margin_top_in   = 0.5,
      margin_left_in  = 0.5,
      margin_right_in = 0.5,
      margin_bottom_in = 0.5
    ),
    default_format = list(
      font_size_half_points = 16L,
      row_height_twips      = 169L
    )
  ) |>
  rtf_section(secinfo = list(
    header = rtf_header(titles),
    footer = rtf_footer(footers))) |>
  rtf_tables(pages)

rtf_path <- file.path(tempdir(), "L_16_2_4_2_1_2.rtf")
generate_rtfreport(rep, rtf_path, overwrite = TRUE)
file.exists(rtf_path)
```

Before writing the file you can eyeball a page in the console -- rtfreporter's
`print()` lays the table out with its headers, rules and alignment:

```{r preview, eval = has_rtf}
print(pages[[1]], n = 8)
```

## Notes

**One set of numbers, not two.** If you are porting hand-written listing code,
you probably have a wrapping width *and* a separate relative column width for
each column, kept in step by eye. `width` replaces both: the rendered columns
come out proportional to the widths you wrapped at, which is almost always
what the two sets were approximating anyway.

**Widths are display widths.** Japanese and other full-width text counts two
columns per glyph, so a `width = 20` column holds ten full-width characters.
Wrapping breaks in the right place without a separate code path.

**Expressions become columns.** `listing_col()` names columns; compute
anything else in a `mutate()` first, as with the rounded platinum-free
intervals above.

**One template for now.** `template = "type1"` is the only one implemented.
The engine underneath -- composition, wrapping, record alignment, spacers,
metadata -- is what a second template would reuse.
