---
title: "A cookbook of clinical listings"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 120)
has_rtf <- requireNamespace("rtfreporter", quietly = TRUE)
```

Almost every listing in a clinical submission is one of three shapes:

| shape | example | what makes it awkward |
|---|---|---|
| **one record per subject** | subject disposition, demographics | wide -- many short columns competing for one page |
| **many records per subject** | adverse events, concomitant medications | the subject must print once, not on every event |
| **subject x parameter x visit** | laboratory results, vital signs | *two* keys to suppress, and they nest |

This article builds one of each, end to end, and leaves you a rendered `.rtf`
to open in Word. The three are deliberately not toy examples: the data has
missing end dates, a controlled term with a `/` in it, and a parameter name
longer than its column -- the things that actually go wrong.

If you have not met `rtf_listing()` before, read
[Building a clinical listing](clinical-listing.html) first; it explains the
engine. This one is about using it.

```{r setup, message = FALSE}
library(ydisctools)
library(rtfreporter)
```

```{r asset-dir, include = FALSE}
# Rendered listings are written where pkgdown publishes them, so the download
# links below resolve on the site.  Outside a repo checkout, fall back to a
# temporary directory so the article still knits.
rtf_dir <- local({
  d <- file.path("..", "..", "pkgdown", "assets", "listings")
  if (dir.exists(file.path("..", "..", "pkgdown"))) {
    dir.create(d, showWarnings = FALSE, recursive = TRUE)
    d
  } else {
    tempdir()
  }
})

# Every listing here shares the same page and the same furniture, so the
# document assembly is written once.
render_listing <- function(pages, file, titles, footers) {
  rtf_document(
    page = list(
      orientation      = "landscape",
      paper_size       = "A4",
      margin_top_in    = 0.6,
      margin_left_in   = 0.5,
      margin_right_in  = 0.5,
      margin_bottom_in = 0.6
    ),
    default_format = list(
      font_size_half_points = 16L,
      row_height_twips      = 169L
    )
  ) |>
    rtf_section(secinfo = list(header = rtf_header(titles),
                               footer = rtf_footer(footers))) |>
    rtf_tables(pages) |>
    generate_rtfreport(file.path(rtf_dir, file), overwrite = TRUE)
  invisible(file)
}
```

All data below is synthetic, shaped after the CDISC pilot ADaM data. Nothing
here is real.

## 1. One record per subject -- subject disposition

The wide, flat case. Eight subjects, one row each, and the difficulty is that
a disposition listing wants a dozen short facts side by side.

```{r adsl-data}
adsl <- data.frame(
  SITEID  = c("701", "701", "701", "703", "703", "708", "708", "716"),
  USUBJID = c("01-701-1015", "01-701-1023", "01-701-1047", "01-703-1086",
              "01-703-1096", "01-708-1187", "01-708-1253", "01-716-1024"),
  ARM     = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose",
              "Placebo", "Xanomeline High Dose", "Xanomeline Low Dose",
              "Placebo", "Xanomeline High Dose"),
  AGE     = c(63L, 64L, 71L, 74L, 77L, 85L, 68L, 81L),
  SEX     = c("F", "M", "F", "M", "F", "F", "M", "F"),
  RACE    = c("WHITE", "WHITE", "BLACK OR AFRICAN AMERICAN", "WHITE", "WHITE",
              "AMERICAN INDIAN OR ALASKA NATIVE", "WHITE", "WHITE"),
  TRTSDT  = c("2013-02-12", "2013-03-05", "2013-04-18", "2013-05-02",
              "2013-06-14", "2013-07-01", "2013-08-09", "2013-09-23"),
  TRTEDT  = c("2013-08-11", "2013-04-02", "2013-10-16", "2013-05-29",
              "2013-12-11", "2013-07-19", "2014-02-04", "2013-10-01"),
  TRTDURD = c(181L, 29L, 182L, 28L, 181L, 19L, 180L, 9L),
  EOSSTT  = c("COMPLETED", "DISCONTINUED", "COMPLETED", "DISCONTINUED",
              "COMPLETED", "DISCONTINUED", "COMPLETED", "DISCONTINUED"),
  DCSREAS = c(NA, "ADVERSE EVENT", NA, "WITHDRAWAL BY SUBJECT", NA,
              "ADVERSE EVENT", NA, "LACK OF EFFICACY"),
  stringsAsFactors = FALSE
)
str(adsl, vec.len = 2, give.attr = FALSE)
```

### Grouping the columns

Twelve source columns will not fit across a page as twelve display columns.
The move is to **group the ones that are read together** and let them share a
column: site with subject, the demographics together, the treatment dates
together, the disposition with its reason.

Two of those groups want opposite layouts, which is why `layout` is set per
column:

* `AGE`, `SEX`, `RACE` are short and belong on **one line** -- `"flow"` gives
  `63/F/WHITE`;
* the treatment dates are long and belong **stacked**, one per line, so the
  column reads down.

```{r adsl-widths}
adsl_cols <- list(
  listing_col(SITEID, USUBJID,
              header = "Site/\nSubject",                             width = 13),
  listing_col(ARM,
              header = "Treatment Group",                            width = 14),
  listing_col(AGE, SEX, RACE,
              header = "Age/Sex/Race",         layout = "flow",      width = 20),
  listing_col(TRTSDT, TRTEDT, TRTDURD,
              header = "First Dose/\nLast Dose/\nDuration (days)",   width = 14),
  listing_col(EOSSTT, DCSREAS,
              header = "End of Study/\nReason",                      width = 22)
)

do.call(auto_listing_widths, c(list(adsl), adsl_cols, list(total_width = 90)))
```

`auto_listing_widths()` reports each column's measured demand next to the width
it proposes. Read it as a diagnosis, not a prescription: `AGE_SEX_RACE` demands
32 characters because one subject's race is `BLACK OR AFRICAN AMERICAN`, and
letting that one value dictate the layout would be wrong. The widths above keep
it at 20 and accept that it wraps for two subjects.

### Building and rendering

```{r adsl-build}
lst_adsl <- do.call(rtf_listing, c(list(adsl), adsl_cols))
pages_adsl <- listing_to_rtftables(lst_adsl, max_rows = 24)
length(pages_adsl)
```

```{r adsl-preview, eval = has_rtf}
print(pages_adsl[[1]])
```

Note what `DCSREAS = NA` did: subject `01-701-1015` completed the study, so
there is no reason -- and the cell reads `COMPLETED`, not `COMPLETED/`. Empty
components are dropped when the column is composed, so a missing value never
leaves a dangling separator behind.

```{r adsl-render, eval = has_rtf}
render_listing(
  pages_adsl, "L_16_2_1_disposition.rtf",
  titles  = list("Listing 16.2.1",
                 "Subject Disposition",
                 "<All Randomized Subjects>"),
  footers = list(c(l = "Duration = last dose - first dose + 1."),
                 c(l = "Note: synthetic data, for illustration only."))
)
```

**[Download `L_16_2_1_disposition.rtf`](../listings/L_16_2_1_disposition.rtf)**

## 2. Many records per subject -- adverse events

Now the subject repeats. Ten events across four subjects, and the listing must
print each subject once, at the head of its run.

```{r adae-data}
adae <- data.frame(
  USUBJID = c(rep("01-701-1023", 3), rep("01-703-1086", 2),
              rep("01-708-1187", 4), "01-716-1024"),
  AESEQ   = c(1L, 2L, 3L, 1L, 2L, 1L, 2L, 3L, 4L, 1L),
  AEBODSYS = c("GASTROINTESTINAL DISORDERS",
               "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
               "NERVOUS SYSTEM DISORDERS",
               "CARDIAC DISORDERS",
               "GENERAL DISORDERS AND ADMINISTRATION SITE CONDITIONS",
               "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
               "SKIN AND SUBCUTANEOUS TISSUE DISORDERS",
               "NERVOUS SYSTEM DISORDERS",
               "INFECTIONS AND INFESTATIONS",
               "PSYCHIATRIC DISORDERS"),
  AEDECOD = c("Diarrhoea", "Erythema", "Dizziness", "Atrial fibrillation",
              "Fatigue", "Application site pruritus", "Rash", "Headache",
              "Upper respiratory tract infection", "Insomnia"),
  AETERM  = c("LOOSE STOOLS", "REDNESS AT PATCH SITE", "LIGHTHEADED",
              "IRREGULAR HEARTBEAT", "TIREDNESS", "ITCHING UNDER PATCH",
              "SKIN RASH ON FOREARM", "HEADACHE", "HEAD COLD",
              "TROUBLE SLEEPING"),
  ASTDT   = c("2013-03-08", "2013-03-19", "2013-03-27", "2013-05-05",
              "2013-05-18", "2013-07-03", "2013-07-05", "2013-07-11",
              "2013-07-14", "2013-09-26"),
  AENDT   = c("2013-03-12", "2013-04-02", NA, "2013-05-11", "2013-05-29",
              "2013-07-08", "2013-07-19", "2013-07-13", "2013-07-19",
              "2013-09-30"),
  ASTDY   = c(4L, 15L, 23L, 4L, 17L, 3L, 5L, 11L, 14L, 4L),
  AESEV   = c("MILD", "MODERATE", "MILD", "SEVERE", "MODERATE", "MILD",
              "MODERATE", "MILD", "MILD", "MODERATE"),
  AESER   = c("N", "N", "N", "Y", "N", "N", "N", "N", "N", "N"),
  AEREL   = c("POSSIBLE", "PROBABLE", "NOT RELATED", "NOT RELATED",
              "POSSIBLE", "PROBABLE", "PROBABLE", "NOT RELATED",
              "NOT RELATED", "POSSIBLE"),
  AEACN   = c("DOSE NOT CHANGED", "DRUG WITHDRAWN", "DOSE NOT CHANGED",
              "DRUG WITHDRAWN", "DOSE NOT CHANGED", "DOSE NOT CHANGED",
              "DRUG WITHDRAWN", "DOSE NOT CHANGED", "DOSE NOT CHANGED",
              "DOSE NOT CHANGED"),
  AEOUT   = c("RECOVERED/RESOLVED", "RECOVERED/RESOLVED", "NOT RECOVERED",
              "RECOVERED/RESOLVED", "RECOVERED/RESOLVED", "RECOVERED/RESOLVED",
              "RECOVERED/RESOLVED", "RECOVERED/RESOLVED", "RECOVERED/RESOLVED",
              "RECOVERED/RESOLVED"),
  stringsAsFactors = FALSE
)
```

### Two things to fix before building

**A missing end date must say so.** Event 3 of subject `01-701-1023` had not
resolved. Left as `NA` it is simply dropped, and the cell reads as if the study
day were the end date. A listing has to say *ongoing*, and since
`listing_col()` takes column names rather than expressions, that becomes a
column first:

```{r adae-derive}
adae$AENDT_C <- ifelse(is.na(adae$AENDT), "ONGOING", adae$AENDT)
```

**A separator must not occur in the data.** `AEOUT` is the CDISC term
`RECOVERED/RESOLVED` -- which contains the default separator. Joined with `/`,
the column would break the controlled term in half:

```{r adae-sep-problem}
listing_wrap("DOSE NOT CHANGED/RECOVERED/RESOLVED", 20)
```

Three lines, and `RECOVERED/RESOLVED` has been split into two of them. Choosing
a separator the data cannot contain fixes it:

```{r adae-sep-fixed}
listing_wrap("DOSE NOT CHANGED / RECOVERED/RESOLVED", 20, sep = " / ")
```

### Building and rendering

`collapse_repeats = TRUE` on the subject column is the whole point of this
listing:

```{r adae-build}
lst_ae <- rtf_listing(
  adae,
  listing_col(USUBJID,
              header = "Subject", collapse_repeats = TRUE,      width = 12),
  listing_col(AESEQ,
              header = "AE\nNo.",                               width = 4),
  listing_col(AEBODSYS, AEDECOD,
              header = "System Organ Class/\nPreferred Term",   width = 26),
  listing_col(AETERM,
              header = "Reported Term",                         width = 18),
  listing_col(ASTDT, AENDT_C, ASTDY,
              header = "Onset/\nResolved/\nStudy Day",          width = 12),
  listing_col(AESEV, AESER, AEREL,
              header = "Severity/Serious/\nRelationship",
              layout = "flow",                                  width = 20),
  listing_col(AEACN, AEOUT,
              header = "Action Taken/\nOutcome", sep = " / ",   width = 22)
)
pages_ae <- listing_to_rtftables(lst_ae, max_rows = 26)
```

```{r adae-preview, eval = has_rtf}
print(pages_ae[[1]])
```

The subject prints once per run, the SOC/PT pair stacks and wraps, and the
severity triple flows onto one line.

Now the part that matters for a regulated deliverable -- **the subject is
printed again at the top of the next page**, not inherited from the page
before:

```{r adae-page2, eval = has_rtf}
print(pages_ae[[2]])
```

That works because `rtf_listing()` does not blank the value itself. It carries
the subject down its record and lets `as_rtftables(collapse_repeats = )` do the
suppression, and rtfreporter suppresses **per page, after the split**. Blanking
at composition time could not know where the page breaks fall.

```{r adae-render, eval = has_rtf}
render_listing(
  pages_ae, "L_16_2_7_adverse_events.rtf",
  titles  = list("Listing 16.2.7",
                 "Adverse Events",
                 "<Safety Analysis Set>"),
  footers = list(c(l = "Serious: Y = yes, N = no.  Study day relative to first dose."),
                 c(l = "MedDRA version 25.0.  Synthetic data, for illustration only."))
)
```

**[Download `L_16_2_7_adverse_events.rtf`](../listings/L_16_2_7_adverse_events.rtf)**

## 3. Subject x parameter x visit -- laboratory results

Two keys now, and they nest: within a subject the parameter repeats, and when
the subject changes the parameter must print again even if it happens to be the
same one.

```{r adlb-data}
lb_block <- function(subj, param, unit, lo, hi, visits, vals, dates) {
  data.frame(USUBJID = subj, PARAM = param, AVALU = unit,
             ANRLO = lo, ANRHI = hi, AVISIT = visits, AVAL = vals, ADT = dates,
             stringsAsFactors = FALSE)
}

adlb <- rbind(
  lb_block("01-701-1015", "Alanine Aminotransferase", "U/L", "10", "40",
           c("Baseline", "Week 4", "Week 12"), c("22", "31", "58"),
           c("2013-02-12", "2013-03-13", "2013-05-08")),
  lb_block("01-701-1015", "Hemoglobin", "g/dL", "12.0", "16.0",
           c("Baseline", "Week 4", "Week 12"), c("13.4", "12.9", "11.6"),
           c("2013-02-12", "2013-03-13", "2013-05-08")),
  lb_block("01-703-1086", "Alanine Aminotransferase", "U/L", "10", "40",
           c("Baseline", "Week 4"), c("35", "41"),
           c("2013-05-02", "2013-05-29")),
  lb_block("01-703-1086", "Platelets", "10^9/L", "150", "400",
           c("Baseline", "Week 4"), c("212", "148"),
           c("2013-05-02", "2013-05-29"))
)
adlb$ANRIND <- c("NORMAL", "NORMAL", "HIGH", "NORMAL", "NORMAL", "LOW",
                 "NORMAL", "HIGH", "NORMAL", "LOW")
```

### Suppression is hierarchical, in declaration order

Mark both keys. Because the subject column is declared first, a change in
subject restarts the parameter's run -- which is exactly what you want, and
what you would otherwise have to code by hand.

Two separators earn their keep here: the result joins its unit with a **space**
(`22 U/L`, not `22/U/L`), and the reference range joins its bounds with
`" - "` (`10 - 40`).

```{r adlb-build}
lst_lb <- rtf_listing(
  adlb,
  listing_col(USUBJID,
              header = "Subject", collapse_repeats = TRUE,    width = 12),
  listing_col(PARAM,
              header = "Parameter", collapse_repeats = TRUE,  width = 26),
  listing_col(AVISIT, ADT,
              header = "Visit/\nDate",                        width = 12),
  listing_col(AVAL, AVALU,
              header = "Result", sep = " ", layout = "flow",  width = 12),
  listing_col(ANRLO, ANRHI,
              header = "Reference\nRange", sep = " - ", layout = "flow",
                                                              width = 13),
  listing_col(ANRIND,
              header = "Flag",                                width = 8)
)
pages_lb <- listing_to_rtftables(lst_lb, max_rows = 30)
```

```{r adlb-preview, eval = has_rtf}
print(pages_lb[[1]])
```

### Why `Parameter` is 26 characters wide

`Alanine Aminotransferase` is 24 characters. At `width = 16` it would wrap onto
two lines -- and **a cell that wraps is never suppressed**. It is printed in
full instead, which is never wrong but reads badly when the same two-line name
repeats down the page:

```{r adlb-too-narrow, eval = has_rtf}
narrow <- rtf_listing(
  adlb,
  listing_col(USUBJID, header = "Subject", collapse_repeats = TRUE, width = 12),
  listing_col(PARAM,   header = "Parameter", collapse_repeats = TRUE, width = 16),
  listing_col(AVISIT,  header = "Visit",                            width = 10),
  listing_col(AVAL,    header = "Result",                           width = 8)
)
print(listing_to_rtftables(narrow)[[1]])
```

So the rule is simple: **give a `collapse_repeats` column enough width to hold
its longest value on one line.** Repeat suppression is for short key columns.

```{r adlb-render, eval = has_rtf}
render_listing(
  pages_lb, "L_16_2_8_laboratory.rtf",
  titles  = list("Listing 16.2.8",
                 "Laboratory Results",
                 "<Safety Analysis Set>"),
  footers = list(c(l = "Flag: HIGH / LOW relative to the reference range."),
                 c(l = "Note: synthetic data, for illustration only."))
)
```

**[Download `L_16_2_8_laboratory.rtf`](../listings/L_16_2_8_laboratory.rtf)**

## What each shape needed

```{r recap, echo = FALSE}
knitr::kable(data.frame(
  Setting = c("`collapse_repeats`", "`layout = \"flow\"`", "custom `sep`",
              "derived column first", "`max_rows`"),
  Disposition = c("--", "demographics", "--", "--", "yes"),
  `Adverse events` = c("subject", "severity triple", "`\" / \"` (outcome has a `/`)",
                       "`ONGOING` end date", "yes"),
  Laboratory = c("subject + parameter", "result, range",
                 "`\" \"`, `\" - \"`", "--", "yes"),
  check.names = FALSE
))
```

Everything else -- composition, wrapping, record alignment, spacer columns,
the hidden record key, pagination that never cuts a record -- is the same in
all three, and none of it appears in the code above.

## Five things that bite

**Pick a `sep` your data cannot contain.** `RECOVERED/RESOLVED`,
`RECOVERED/RESOLVED WITH SEQUELAE` and `10^9/L` all carry the default
separator. `" / "` and `" - "` are usually safe because the values do not
contain spaces around their slashes.

**A `collapse_repeats` column must fit on one line.** A wrapped cell is printed
in full rather than suppressed. Widen the column, or shorten the value.

**`listing_col()` names columns, it does not evaluate expressions.** Rounding,
formatting, `ONGOING`, a derived flag -- all of it becomes a column with
`mutate()` or `$<-` before the listing is built.

**Missing is not the same as blank.** An `NA` component is dropped silently,
which is right for "this subject has no discontinuation reason" and wrong for
"this event has not resolved". Only you know which one it is.

**Widths are display widths, and `auto_listing_widths()` is a diagnosis.** It
reports what the data demands; the demand of a single long value is information,
not an instruction. Full-width Japanese glyphs count as two, so a `width = 20`
column holds ten of them.

## The rendered files

```{r files, echo = FALSE, eval = has_rtf}
knitr::kable(data.frame(
  Listing = c("16.2.1 Subject Disposition",
              "16.2.7 Adverse Events",
              "16.2.8 Laboratory Results"),
  Shape = c("one record per subject",
            "many records per subject",
            "subject x parameter x visit"),
  File = c("[L_16_2_1_disposition.rtf](../listings/L_16_2_1_disposition.rtf)",
           "[L_16_2_7_adverse_events.rtf](../listings/L_16_2_7_adverse_events.rtf)",
           "[L_16_2_8_laboratory.rtf](../listings/L_16_2_8_laboratory.rtf)")
))
```

Open them in Word: A4 landscape, 8 pt Courier, a title block and footnotes on
every page, and -- in the adverse-events listing -- a subject reprinted at the
top of page 2.
