---
title: "From the SAP to the tables: TOC, shells, ARS and the table programme"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
```

This article walks the **whole ydisctools ARS toolchain** on a bundled
synthetic study -- STUDY01, a decent-scale sample SAP with ten planned
displays -- from the source Word documents down to the final RTF tables:

```
company CSR template ----> read_csr_map()  ..... numbering section map
study SAP ---------------> read_sap_toc()  ..... draft TOC rows   (DRAFT)
        |
        v  human / agent review  (the semi-automated step, materialised below)
completed TOC workbook  ==== the hub, single source of truth ====
        |                        |
        v                        v
   ars_mock()               ars_from_toc() |> build_ars() |> write_ars_xlsx()
   (mock shells)                 |
                                 v
                    siera::readARS() -> ARD programmes -> combined study ARD
                                 |
                                 v
                    table programme  (format_stats() + pivot_stats_wider()
                                 |    + rtfreporter)
                                 v
                          final tables (RTF)
```

## 0. Every stage's output ships with the package

You do not have to execute anything to inspect the chain: the installed
package bundles the **artefact of every stage** in `inst/sap-pipeline/`,
regenerated end-to-end by one script (`make_sap_pipeline.R`):

```{r setup, message = FALSE}
library(ydisctools)
pipe <- system.file("sap-pipeline", package = "ydisctools")
list.files(pipe, recursive = TRUE)
```

`adam/` is the input (deterministic dummy ADaM for STUDY01 -- 60 subjects,
three CDISC-pilot-flavoured arms), `01_toc` through `06_tfl` are the six
stage outputs.  Open `06_tfl/TFL_STUDY01.rtf` in Word to see where this
article ends up.  The rest of the article *re-runs* the same chain in
`tempdir()` so you can watch each artefact being made; every section notes
the bundled file it corresponds to.

The two source Word documents live next door.  Real SAP/CSR templates
(TransCelerate's suite, SAPs on clinicaltrials.gov) cannot be
redistributed, so ydisctools bundles two **original synthetic samples**
(only the section layout is informed by the public ICH E3 guideline; every
sentence is our own -- see the folder README for the provenance note):

```{r}
samples <- system.file("ars-doc-samples", package = "ydisctools")
list.files(samples)
```

Two deliberate demo twists: the company template files its displays under
section **15** (not ICH E3's 14), and the SAP's planned-display appendix
leaves the display numbers **blank** -- numbering is the TFL spec's job.

## 1. Numbering map from the company CSR template

`read_csr_map()` walks the template's headings, anchors itself to the
tables section, and keyword-maps the subsections to the numbering keys.
Literal heading numbers are used when present; templates that rely on Word
auto-numbering are reconstructed rule-based from the heading levels (and
flagged for review):

```{r}
csr <- read_csr_map(file.path(samples, "sample_csr_template.docx"))
csr$section_map
csr$notes
```

The company convention (15.x) will now drive every display number -- no
ICH default leaks through.

## 2. Draft TOC from the SAP

`read_sap_toc()` finds the planned-display table in the SAP appendix and
drafts one TOC row per display: titles are keyword-matched onto the bundled
display recipes, everything undecidable becomes a review note.  (SAPs whose
display list lives in the Statistical Methods *prose* instead go through
the prose route -- see `?read_sap_toc`.)

```{r}
draft <- read_sap_toc(file.path(samples, "sample_sap.docx"))
draft$toc[, c("title", "display_type", "population")]
draft$notes
```

Ten displays: demographics, disposition and exposure; five adverse-event
views -- note row 8, the **serious**-TEAE display, matched the same
`ae_soc` recipe as row 5; and two `custom` efficacy displays the recipes
cannot know about.

## 3. The review step (what the human -- or agent -- edits)

This is the deliberately semi-automated phase.  The draft needs decisions
that live in the SAP's *prose* (its Analysis Sets, endpoint and safety
sections), applied here as explicit edits so the chain continues -- do the
same in Excel if you prefer:

```{r}
toc <- draft$toc

# (a) meaningful output ids
toc$output_id <- c("Out_dm", "Out_ds", "Out_ex",
                   "Out_ae_ov", "Out_ae_soc", "Out_ae_pt", "Out_ae_sev",
                   "Out_ae_ser", "Out_eff_pri", "Out_eff_sec")

# (b) populations: the SAP defines Safety = ADSL.SAFFL, ITT = ADSL.ITTFL
toc$population <- ifelse(toc$population == "ITT", "ITTFL", "SAFFL")

# (c) arm grouping and the three arms named in SAP section 1
toc$group_by <- "TRT01A"
toc$groups   <- "Placebo | Xanomeline Low Dose | Xanomeline High Dose"

# (d) the serious-AE display is the ae_soc recipe narrowed to AESER = 'Y';
#     the extra condition applies to every analysis EXCEPT the big-N
#     denominator, which stays at analysis-set level
toc$where[toc$output_id == "Out_ae_ser"] <- "AESER EQ Y"
```

The two custom efficacy displays need hand-authored analyses (the SAP's
section 2 defines the endpoints as `ADEFF` parameters `PRIMEP` / `SECEP`
with responder flag `CRIT1FL`):

```{r}
eff_analyses <- function(output_id, paramcd) {
  data.frame(
    output_id   = output_id,
    analysis_id = "",
    name        = c("Number of subjects",
                    "Change from baseline at Week 24",
                    "Responders, n (%)"),
    method      = c("total_n", "continuous_summary", "categorical_summary"),
    dataset     = c("ADSL", "ADEFF", "ADEFF"),
    variable    = c("USUBJID", "AVAL", "USUBJID"),
    population  = "",
    group_by    = c("", "", "TRT01A, CRIT1FL"),
    groups      = "", groups2 = "",
    where       = c("", paste("PARAMCD EQ", paramcd),
                    paste("PARAMCD EQ", paramcd)),
    denominator = c("", "", "auto"),
    stringsAsFactors = FALSE
  )
}
eff <- rbind(eff_analyses("Out_eff_pri", "PRIMEP"),
             eff_analyses("Out_eff_sec", "SECEP"))
```

Write the completed TOC workbook -- the reviewed artefact every later phase
consumes, and the natural thing to keep under version control:

```{r}
toc_file <- file.path(tempdir(), "TOC_STUDY01.xlsx")
writexl::write_xlsx(
  list(Study = data.frame(
         key = c("study_id", "study_title"),
         value = c("STUDY01", "A Randomised Study of Xanomeline in Dementia")),
       TOC = toc, Analyses = eff),
  toc_file
)
```

*Bundled artefact:* `sap-pipeline/01_toc/TOC_STUDY01.xlsx` -- this same
workbook, written by the pipeline script.

## 4. Mock shells

Shells derive from the same TOC (and the same recipes) as the ARS, so they
cannot drift apart.  One sheet per display; the sponsor numbering from
step 1 lands in the title block:

```{r}
mock_file <- file.path(tempdir(), "shells_STUDY01.xlsx")
ars_mock(toc_file, mock_file, section_map = csr$section_map)
head(as.data.frame(suppressMessages(
  readxl::read_excel(mock_file, sheet = "Out_dm", col_names = FALSE))), 10)
```

`format = "rtf"` renders the same shells through
[rtfreporter](https://ichirio.github.io/rtfreporter/) instead -- one page
per display, with the furniture the final tables will use.

*Bundled artefacts:* `sap-pipeline/02_shells/shells_STUDY01.xlsx` and
`.rtf` (all ten displays in both formats).

## 5. ARS metadata

The same TOC + section map expand into the full ARS workbook.  Note the
auto-assigned numbers following the *company* convention -- baseline
displays under 15.1, efficacy under 15.2, adverse events under 15.3.1:

```{r}
params <- ars_from_toc(toc_file, section_map = csr$section_map)
params$toc[, c("toc_no", "output_id", "display_type")]

ars <- build_ars(params)
ars_file <- file.path(tempdir(), "ARS_STUDY01.xlsx")
write_ars_xlsx(ars, ars_file, overwrite = TRUE)
```

*Bundled artefact:* `sap-pipeline/03_ars/ARS_STUDY01.xlsx` (nine
interlinked, siera-ready sheets).

## 6. ARD programmes -- and one combined study ARD

`ars_generate_ard()` hands the workbook to `siera::readARS()` (one
generated {cards} programme per display, kept for inspection), runs each
programme against the ADaM folder, and binds everything into one ARD keyed
by `OutputId`.  The bundled dummy data carries every variable the ten
displays need, so **the whole study runs**:

```{r, message = FALSE, warning = FALSE}
prog_dir <- file.path(tempdir(), "ard_programmes")
ard <- ars_generate_ard(ars_file, adam_path = file.path(pipe, "adam"),
                        output_path = prog_dir)
basename(attr(ard, "scripts"))
table(ard$OutputId)
```

A taste of the result -- age-group counts from the demographics display:

```{r}
ard[ard$AnalysisId == "An_03" & ard$stat_name %in% c("n", "p"),
    c("OutputId", "group1_level", "variable", "variable_level",
      "stat_name", "stat")] |> head(6)
```

*Bundled artefacts:* `sap-pipeline/04_programmes/ARD_Out_*.R` (the ten
programmes; run them from the pipeline folder) and
`sap-pipeline/05_ard/ARD_STUDY01.csv` (the combined ARD, flattened for
inspection).

## 7. The table programme

The last leg is a **study table programme**: ARD numbers in, submission
tables out.  Two small ydisctools helpers do the shaping --
`format_stats()` collapses the statistic rows into display strings via a
template table, `pivot_stats_wider()` spreads the arms into columns.
Here is the whole pattern on the Age analysis:

```{r, message = FALSE}
age <- ard[ard$AnalysisId == "An_02", ]
lbl <- c(N = "n", mean = "Mean (SD)", sd = "Mean (SD)", median = "Median",
         p25 = "Q1, Q3", p75 = "Q1, Q3", min = "Min, Max", max = "Min, Max")
long <- data.frame(arm   = age$group1_level,
                   label = unname(lbl[age$stat_name]),
                   stat  = age$stat_name,
                   value = unlist(age$stat))
long <- long[!is.na(long$label), ]

specs <- data.frame(
  label    = c("n", "Mean (SD)", "Median", "Q1, Q3", "Min, Max"),
  template = c("{N:0}", "{mean:1} ({sd:2})", "{median:1}",
               "{p25:1}, {p75:1}", "{min:0}, {max:0}"))

format_stats(long, specs, stat_col = "stat", value_col = "value") |>
  pivot_stats_wider(expand_cols = "arm")
```

The bundled table programme, `sap-pipeline/06_tfl/make_tfl.R`,
generalises exactly this: three renderers (big-N header, continuous block,
categorical block) driven by the TOC's analysis metadata, then
[rtfreporter](https://ichirio.github.io/rtfreporter/) writes one landscape
page per display:

```{r, eval = FALSE}
# from the pipeline folder (inst/sap-pipeline):
#     Rscript 06_tfl/make_tfl.R
# reads  01_toc/TOC_STUDY01.xlsx + 05_ard/ARD_STUDY01.csv
# writes 06_tfl/TFL_STUDY01.rtf   -- all ten displays, one per page
```

*Bundled artefact:* `sap-pipeline/06_tfl/TFL_STUDY01.rtf` -- the finished
document, next to the programme that made it.

## Where to next

[The ARS chain](ars-chain.html) looks at the middle of this pipeline in
close-up -- the compact parameter workbook, `build_ars()`'s expansion
rules, and recovering parameters from an *existing* programme with
`ars_params_from_code()` when you start from a past study instead of a
SAP.
