---
title: "From the SAP to the ARS: TOC, shells and ARD programmes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
```

This article walks the **upstream half** of the ydisctools ARS toolchain on
two bundled sample documents:

```
company CSR template ----> read_csr_map()  ..... numbering section map
study SAP ---------------> read_sap_toc()  ..... draft TOC rows   (DRAFT)
        |
        v  human / agent review  (the semi-automated step, materialised below)
completed TOC workbook  ==== the hub, single source of truth ====
        |                        |
        v                        v
   ars_mock()               ars_from_toc() |> build_ars() |> write_ars_xlsx()
   (shells)                      |
                                 v
                    siera::readARS() -> ARD programmes
```

The downstream half -- running the generated programmes into one
`OutputId`-keyed study ARD -- is covered by
[The ARS chain](ars-chain.html); this article ends where that one begins.

## 0. The bundled sample documents

Real SAP/CSR templates (TransCelerate's suite, SAPs on clinicaltrials.gov)
cannot be redistributed, so ydisctools bundles two **original synthetic
samples** (only the section layout is informed by the public ICH E3
guideline; every sentence is our own -- see the folder README for the
provenance note):

```{r setup, message = FALSE}
library(ydisctools)
samples <- system.file("ars-doc-samples", package = "ydisctools")
list.files(samples)
```

Two deliberate demo twists: the company template files its displays under
section **15** (not ICH E3's 14), and the SAP's planned-display appendix
leaves the display numbers **blank** -- numbering is the TFL spec's job.

## 1. Numbering map from the company CSR template

`read_csr_map()` walks the template's headings, anchors itself to the
tables section, and keyword-maps the subsections to the numbering keys.
Literal heading numbers are used when present; templates that rely on Word
auto-numbering are reconstructed rule-based from the heading levels (and
flagged for review):

```{r}
csr <- read_csr_map(file.path(samples, "sample_csr_template.docx"))
csr$section_map
csr$notes
```

The company convention (15.x) will now drive every display number -- no
ICH default leaks through.

## 2. Draft TOC from the SAP

`read_sap_toc()` finds the planned-display table in the SAP and drafts TOC
rows: titles are keyword-matched onto the bundled display recipes,
everything undecidable becomes a review note:

```{r}
draft <- read_sap_toc(file.path(samples, "sample_sap.docx"))
draft$toc[, c("toc_no", "output_id", "title", "display_type", "population")]
draft$notes
```

## 3. The review step (what the human -- or agent -- edits)

This is the deliberately semi-automated phase. The draft needs four
decisions that live in the SAP's *prose* (its Analysis Sets and design
sections), and here they are applied as explicit edits so the chain can
continue -- do the same in Excel if you prefer:

```{r}
toc <- draft$toc

# (a) meaningful output ids
toc$output_id <- c("Out_dm", "Out_ds", "Out_ex", "Out_ae_ov",
                   "Out_ae_soc", "Out_ae_sev", "Out_eff")

# (b) populations: the SAP defines Safety = ADSL.SAFFL, ITT = ADSL.ITTFL
toc$population <- ifelse(toc$population == "ITT", "ITTFL", "SAFFL")

# (c) arm grouping and the three arms named in SAP section 1
toc$group_by <- "TRT01A"
toc$groups   <- "Placebo | Xanomeline Low Dose | Xanomeline High Dose"

# (d) the efficacy display is 'custom': author its analyses by hand
eff <- data.frame(
  output_id   = "Out_eff",
  analysis_id = "",
  name        = c("Number of subjects", "Primary endpoint",
                  "Responders, n (%)"),
  method      = c("total_n", "continuous_summary", "categorical_summary"),
  dataset     = c("ADSL", "ADEFF", "ADEFF"),
  variable    = c("USUBJID", "AVAL", "USUBJID"),
  population  = "", group_by = c("", "", "TRT01A, CRIT1FL"),
  groups = "", groups2 = "",
  where       = c("", "PARAMCD EQ PRIMEP", "PARAMCD EQ PRIMEP"),
  denominator = c("", "", "auto"),
  stringsAsFactors = FALSE
)
```

Write the completed TOC workbook -- the reviewed artefact every later phase
consumes (and the natural thing to keep under version control):

```{r}
toc_file <- file.path(tempdir(), "TOC_STUDY01.xlsx")
writexl::write_xlsx(
  list(Study = data.frame(key = c("study_id", "study_title"),
                          value = c("STUDY01", "Sample Study")),
       TOC = toc, Analyses = eff),
  toc_file
)
```

## 4. Mock shells

Shells derive from the same TOC (and the same recipes) as the ARS, so they
cannot drift apart. One sheet per display; the sponsor numbering from step 1
lands in the title block:

```{r}
mock_file <- file.path(tempdir(), "shells_STUDY01.xlsx")
ars_mock(toc_file, mock_file, section_map = csr$section_map)
head(as.data.frame(suppressMessages(
  readxl::read_excel(mock_file, sheet = "Out_dm", col_names = FALSE))), 10)
```

(`format = "rtf"` renders the same shells through
[rtfreporter](https://ichirio.github.io/rtfreporter/) instead -- one page
per display, with the furniture the final tables will use.)

## 5. ARS metadata

The same TOC + section map expand into the full ARS workbook. Note the
auto-assigned numbers following the *company* convention:

```{r}
params <- ars_from_toc(toc_file, section_map = csr$section_map)
params$toc[, c("toc_no", "output_id", "display_type")]

ars <- build_ars(params)
ars_file <- file.path(tempdir(), "ARS_STUDY01.xlsx")
write_ars_xlsx(ars, ars_file, overwrite = TRUE)
```

## 6. ARD programmes

`siera::readARS()` turns the workbook into one runnable {cards} programme
per display:

```{r, message = FALSE}
prog_dir <- file.path(tempdir(), "ard_programmes")
dir.create(prog_dir, showWarnings = FALSE)
siera::readARS(ars_file, output_path = prog_dir, adam_path = "adam")
list.files(prog_dir)
```

And they run: here the demographics display executes against the dummy ADaM
data shipped with the package (the other displays need ADaM variables the
dummy data does not carry -- point `adam_path` at your study data):

```{r, message = FALSE, warning = FALSE}
ex <- system.file("ars-examples", package = "ydisctools")
ard <- ars_generate_ard(ars_file, adam_path = file.path(ex, "adam"),
                        spec_output = "Out_dm")
ard[ard$stat_name %in% c("n", "p") & !is.na(ard$variable_level),
    c("OutputId", "group1_level", "variable", "variable_level",
      "stat_name", "stat")] |> head(6)
```

## Where to next

[The ARS chain](ars-chain.html) continues from exactly this point: running
every generated programme and combining the results into one study-level
ARD keyed by `OutputId`, plus recovering compact parameters from existing
programmes (`ars_params_from_code()`) when you start from a past study
instead of a SAP.
